Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-09-19 joint_inf.tex
:tennis:
version 3...

\author[1]{David A. Shaw}
\author[1]{Frederick A. Matsen IV\thanks{Corresponding author. Email: \url{matsen@fredhutch.org}}}
\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
\date{}

@@

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we outline cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
\end{abstract}

@@

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.

@@

Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset. We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader. Time for a corollary?
%das: ok, I reworded it a bit; let me know how you think it sounds.
\begin{restatable}{theorem}{topoInconsist}
\label{thm:main_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Can we say "open set" or "set of positive measure"?
There exists an open set of $0 < x^*, y^* < 1$ such that for some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:incons-analytic}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.

%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
%das: good idea; I think this will also appease reviewer 1.
%EM On the other hand we should still refer to it in the text around here.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states.
This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Data generated as in Fig.~\ref{fig:incons-analytic}.
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

We provide a full description of our optimization procedure in the Appendix, but briefly, we analytically reduce the general case to 81 (see Table~\ref{tab:likelihoods}) and perform maximizations for each.
For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
Inference on the integrated likelihood performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
Moreover, the errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
%das: yeah, it was kind of hand-wavy.
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
%EM Is this next sentence necessary given the previous one?
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
For the case of estimating branch lengths, we are able to provide empirical estimates for regions of overestimation that are tighter than the analytical bounds, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
Insight from the reviewers, the editors, and Vu Dinh greatly improved this manuscript.

>>>>

\author[1]{David A. Shaw}
\author[2]{Vu C. Dinh}
\author[1]{Frederick A. Matsen IV\thanks{Corresponding author. Email: \url{matsen@fredhutch.org}}}
\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
\affil[2]{Department of Mathematical Sciences, University of Delaware\\ Newark, DE, USA}
\date{}

@@

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except in degenerate cases, and we provide extensive empirical results for outlining the consistent bias in this setting.
\end{abstract}

@@

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
We show that, when the topology is known and fixed, joint inference cannot be consistent except in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.

@@

\subsection*{Inconsistent branch length estimation}

When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ any branch length estimate is consistently biased.

\begin{restatable}{theorem}{blInconsist}
\label{thm:bl_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For all $0 < x^*, y^* < 1$, given some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{t} \neq t^*$.
\end{restatable}

%TODO: more discussion here
In words, the joint estimation procedure never recovers the true generating $t^*$ except in cases of zero or infinite branch length.

\subsection*{Convergence to degenerate topology}

Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset. We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader. Time for a corollary?
%das: ok, I reworded it a bit; let me know how you think it sounds.
%EM This is better, but the notation still seems a little funny. Namely, xi is really something that's determined by Y, but the notation seems to hint that it can be specified a priori. The point is that the ML xi has problems, not that there exists a xi that has problems.
%das: ok, edited
\begin{restatable}{theorem}{topoInconsist}
\label{thm:topo_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists an open set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.

%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
%das: good idea; I think this will also appease reviewer 1.
%EM On the other hand we should still refer to it in the text around here.
%das: it's referred to in the first sentence of above paragraph---good?

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the analytically\hyp{}derived region (Fig.~\ref{fig:max-anc-state}).
To verify the regions of inconsistency and obtain a clearer picture of the closed form parameter estimates, we plot the optimal $\hat{w}$ via joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Data generated as in Fig.~\ref{fig:max-anc-state}.
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

In our optimization procedure, we analytically reduce the general case to 81 separate cases (see Table~\ref{tab:likelihoods}) and obtain a closed form maximal value for each.
For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.

Inference on the integrated likelihood performs as expected where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
We use L-BFGS-B when optimizing \eqref{eq:marginal_likelihood}.
The errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
In addition, we have shown there are no nontrivial generating parameters that yield consistent branch length estimates.

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
%das: yeah, it was kind of hand-wavy. edited
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
Insight from the reviewers and the editors greatly improved this manuscript.
